Preventing a Nuclear ISIL: \\Recommendations for United States Policy

### 2015-05-09 draft.tex
Implemented various changes and corrections. This commit was sent to Denia for proofing, and the Executive Summary and Conclusions.

\subsection{??????? ORDERED LIST OF RECOMMENDATIONS HERE ????????}

@@

\subsection{State functions/income}

@@

\subsection{What makes this a different problem than traditional non-proliferation efforts?}

@@

\subsection{ISIL's potential targets (“declaratory policies”)}

@@

This  report is directed towards the immediate response to development of the nuclear capabilities of ISIL, and does not consider the consequences which might result if ISIL were to evolve into a permanent state. This  time scale greatly reduces the possibility that ISIL possesses the infrastructure necessary to produce SNM or to assemble such material into a functioning warhead, unless they capture territory containing existing infrastructure. Due to the extreme sophistication in the triggering device, the development of fusion weapons is very unlikely in the time window considered. Rather, emphasis is placed on diverting existing SNM from existing global stockpiles.

@@

\section{Limitations and Assumptions of this Report}

\subsection{?????? Needs content proofing ?????}

@@

\autoref{app:weapons}  covers the basic categories of weapons enabled by nuclear materials: RDDs, fission weapons, and fusion weapons. This Appendix builds on the concepts outlined in \autoref{app:physics} and \autoref{app:app_fuel_cycle}. Any ISIL pursuit of nuclear weapons or nuclear technology would ultimately result in the development of one of these three types of weapons.

@@

\subsection{Introduction}

@@

\section{Post-Detonation Forensics Methods}

@@

\subsection{Data Processing and Response Time}

@@

\subsection{????????  Update logic tree with Julie's new colored one ??????????}

\includepdf[angle=270, pagecommand={\thispagestyle{fancy2}\null\vfill\captionof{figure}{Non-linear Logic Tree}\label{fig:logic_tree}}, scale=0.85]{./figures/NucSec_Logic_Model.pdf}

@@

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.55]{./figures/Policy_Chart2.png}
   \caption{Summary of existing and proposed policies to combat ISIL acquisition of nuclear weapons.}
     \label{fig:policy_chart}
\end{figure}

@@

A key step for the United States in preventing ISIL acquisition and use of a nuclear device is to coordinate and organize the various initiatives across the government that are handling various pieces of this puzzle. In 2008, the authors of a Belfer Center report on preventing nuclear terrorism recommended that President Obama appoint a senior figure to a post in the National Security Council (suggesting this may be a new Deputy National Security Advisor) with the sole responsibility of waking up every morning and thinking about how to prevent a nuclear terrorist attack \cite{Bronner2008}. Since that time, President Obama created a new White House post called the White House Coordinator for Defense Policy, Countering Weapons of Mass Destruction, and Arms Control. The National Security Staff's former Senior Director for Europe, Liz Sherwood-Randall was promoted to that position \cite{Rogin2013}. While this is a step in the right direction, the arms-control role of this position is likely to consume much of Ms. Sherwood-Randall's time, forcing her to put some anti-nuclear terrorism initiatives on the backburner. For this reason, the authors of this report recommend that the White House:

\begin{itemize}
 \item Break out the countering nuclear terrorism aspect of the role of White House Coordinator for Defense Policy, Countering Weapons of Mass Destruction, and Arms Control, and create a new White House role within the National Security Council that deals exclusively with nuclear terrorism. 
 \item Limit the responsibilities of this role to nuclear terrorism, leaving out nuclear weapons concerns as related to Pakistan, North Korea, and Iran, as well as arms-control. These concerns would dominate much of this person's time, detracting from the necessary focus on preventing nuclear terrorism.
 \item Fill this position with a senior figure who commands respect across cabinet departments, and is capable of speaking on behalf of the president.
\end{itemize}

This person would have full responsibility for coordinating and overseeing counter-nuclear terrorism efforts. Making this role a Deputy National Security Advisor would have the advantage of giving the role the authority to integrate policies across the Department of Defense, the Department of Energy, the State Department, the Intelligence community, Homeland Security, and law enforcement \cite{Bronner2008}. This person would also have a role working with the Office of Management and Budget (OMB) to ensure that all agency budgets are aligned with the strategy they articulate. Additionally, the President can direct OMB to prepare a crosscut of departmental budgets to find all funds associated with preventing nuclear terrorism and put them under a single budget examiner \cite{Bronner2008}.

The above suggestions are consisted with the 2008 report from the Belfer Center. The report also emphasizes the importance of limiting the scope of this proposed role. In particular, this person should not be responsible to Pakistan, Iran, or North Korea, because these issues would take up all of this person's time \cite{Sharp2008}. This is the key problem with the current National Security Council role occupied by Ms. Sherwood-Randall - with arms control and broader counterterrorism efforts under her purview, there is still need for a senior official at this level in the White House dedicated exclusively to preventing nuclear terrorism.

@@

Another White House priority should be to articulate the central importance of combating ISIL efforts to acquire and use nuclear weapons. Creating a high-level Nuclear Security Council position is the beginning of demonstrating counter-nuclear terrorism as a United States priority and broadcasting this commitment to the international community.

@@

In addition to taking the actions outlined above, the United States should launch a global security campaign aimed at building consensus that steps must be taken to reduce the risk of nuclear terrorism, encouraging allies and other nuclear powers to create high-level positions (similar to the proposed Deputy National Security Advisor), and coordinating state efforts to combat nuclear terrorism.

\subsection{??????????? This ref is never used! ???????????}

[4] Charles D. Ferguson, “Preventing Catastrophic Nuclear Terrorism,” Council on Foreign Relations. CSR No. 11, March 2006. \url{http://www.cfr.org/weapons-of-mass-destruction/preventing-catastrophic-nuclear-terrorism/p10067 }

@@

Policy: \textcolor{red}{Target ISIL Income Stream Proposed Policy}

@@

\subsubsection{Restrict ISIL access to international financial systems}

@@

\subsubsection{Target ISIL leaders and supporters with sanctions}

@@

\subsection{Restrict ISIL access to nuclear weapons experts}

@@

Policy: \textcolor{red}{Limiting passports, continual surveillance, and combatting propaganda}

@@

\subsection{Rollback ISIL territorial gains}

@@

Policy: \textcolor{red}{Targets of stretgic significance in support of allies}

@@

\subsection{??????? Shrink down pictures, two to a page ????????}

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.6]{./figures/current_territory.png}
   \caption{Current ISIL Territory \cite{BBC2015a}}
     \label{fig:current_territory}
\end{figure}

@@

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.4]{./figures/oil_reserves.png}
   \caption{Oil Reserves and Ground Transportation Routes \cite{PoliticalCalculations2012}}
     \label{fig:oil_reserves}
\end{figure}

@@

Policy: \textcolor{red}{Second Line of Defense (SLD) Core Program}

@@

\subsection{???? Need list for references in this section  ?????}

In the event that ISIL attempts to acquire SNM, the core of a nuclear weapon, or a complete weapon, it is likely that it will need to be smuggled internationally. It is in the U.S.'s best interest to prevent a nuclear weapon or SNM from entering the controlled territory of ISIL. There are two methods of trafficking SNM or a complete weapon, overland and overseas. Trafficking of a complete weapon over land would be difficult due to the heavy weight of the weapon. However, the SNM is relatively light weight and volume-wise very small, and it is very likely that over land routes would be used. ISIL is extremely proficient at smuggling and makes millions of dollars a day from smuggling everything from oil to humans [5]. Due to this massive illicit trafficking infrastructure already set up, it is unlikely that any effort to intercept these SNM overland would likely succeed. However, prevention of smuggling SNM overseas through the ports is much more feasible.

To prevent the smuggling of nuclear materials, the U.S., NNSA started the Second Line of Defense (SLD) Core Program [1]. This program installs radiation detection equipment at borders, airports, and strategic feeder ports in Russia, former Soviet Union states, and other key countries [1]. While not specific to the ISIL crisis, it does help hinder any attempts for ISIL to smuggle nuclear materials/weapons. To adapt the SLD for use against ISIL, it is proposed to add SLD port monitors at the Port of Lattakia and the Port of Tartous (see \autoref{fig:seaports}), which are the closest ports to ISIL-controlled territory that have container liner services [2].  By adding two more ports to this list, the likelihood of ISIL succeeding in smuggling a nuclear weapon/materials can be dramatically reduced. In general it is estimated to cost \$3.75 million dollars per port and would take one year to implement [6].

The NNSA is the lead agency and would be responsible for developing the strategic direction of the implementation as well as providing project performance and financial oversight. The main contributors would be the Domestic Nuclear Detection Office (DNDO), Pacific Northwest National Laboratory (PNNL), and Bechtel Nevada. The facilities at the DNDO will be used for testing purposes with the support of Bechtel Nevada. PNNL will be the lead organization for providing training to countries participating in the SLD. Bechtel Nevada will be responsible for routine and preventative maintenance for the equipment deployed overseas [4].

\begin{figure}
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.85]{./figures/seaports.png}
   \caption[Location of the five seaports in Syria.]{Location of the five seaports in Syria. Of these only the blue-colored ones (Port of Lattakia and the Port of Tartous) have container liner services [3].}
     \label{fig:seaports}
\end{figure}

@@

@@

@@

Policy: \textcolor{red}{Operation Inherent Resolve}

@@

\subsection{Flow of non-nuclear but necessary technologies}

Flow of non-nuclear but necessary technologies (Sandia components)
Export control regimes

@@

@@

\chapter{Policy Courses of Action if ISIL is successful in developing a nuclear device}

@@

\chapter{???????   Move bulk of info into appendices    ?????????}

ISIL taking over a major hospital/university - this is a major concern for chem/bio, I think, but not sure if there is a nuclear component. I know that for DoD, terrorist control of failed state capabilities (in the form of former/current facilities) for WMD purposes is a top concern.

Agency cooperation to manage and coordinate a crisis. This would include current actions to prepare to respond to a crisis, and might have lessons learned from Fukushima and Ebola in terms of inter-agency action. How do we utilize agencies and individuals without access to top secret information in the event of a loose nuke? There will be a lot of expertise that could help but intelligence community and white house won't want broad knowledge that ISIL may be getting close to nuclear device.

This goes along with planning and preparing for a situation with a considerable lack of uncertainty. We're not just planning what to do, but planning structures that allow us to respond quickly to uncertainty. Looking for a lost nuke or lost nuclear material is like looking for a needle in a haystack. Since there is never a precise enough tool to find it, how do we shrink the haystack? How do we ensure we're leveraging collective knowledge of various agencies to zero in on it and verify if it's ISIL who has it. Even if we're keeping an eye on ISIL, are we developing capabilities for the event of a lost nuke (or whatever), to look and investigate, is there any way this thing is making its way to ISIL?

Utilizing Proliferation Security Initiative to track, interdict, and dispose of sensitive materials. There is some thinking on this in the Cooperative Threat Reduction.

There is significant competition within DoD over ISR (intelligence, surveillance, recognizance). How do we make sure the counterterrorism/WMD office gets enough of it to monitor nuclear materials in countries with ISIL sympathizers, and if we don't get it, what's the back up plan?

Broadly speaking, working this problem without assuming false precision. We don't want to assume we'll know exactly what ISIL has or where things are going, etc.

Going back and working with the maps that I sent you guys and trying to get all of that information about facilities and power grids and oil onto one or a couple maps.

Any other big gaps that need attention.

\chapter{Conclusions and suggestions }

@@

The most substantial effect of an RDD is the spread of fear, panic, economic disruption, and the cost associated with cleaning up a contaminated area. These devices are unlikely to result in as many deaths as a fission or fusion device due to the short duration of radiation exposure and the diffuse distribution of the dispersed materials. \autoref{tab:health_effects} from the U.S. Environmental Protection Agency shows the level of exposure required to produce various health symptoms and the time from acute exposure to symptom onset \cite{USEPA1999}. Department of Homeland Security (DHS) and local law enforcement crisis management plans to evacuate crowds following an RDD detonation and the subsequent radiation cleanup will drastically reduce the number of people who receive radiation doses large enough to cause short term health effects.

@@

A fusion device is a multistage weapon which consists of at least one fission stage and at least one fusion stage. The fission stages are used to drive the fusion reaction, which in turn provides a short, high energy pulse of neutrons which then drives additional fission. This complicated array of triggering devices is used in turn to achieve explosive yields which far exceed a single fission device. \autoref{fig:ulam_device} shows an example of a fusion device.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{./figures/Ulam_device.png}
  \caption{Schematic representation of the design of a fusion weapon \cite{WikimediaCommons2005}.}
  \label{fig:ulam_device}
\end{figure}

@@

Fusion weapons are far more sophisticated than fission weapons, and essentially consists of three fission devices and one fusion device. It is assumed that ISIL is pursuing nuclear weapons in order to commit acts of terrorism. While higher yields will result in larger damage,  optimizing yield does not necessarily enhance the terror factor of a nuclear device. Due to the complexity of the design of multistage fusion devices, it is assumed that ISIL will elect to pursue  fission weapons, most probably a gun-type design.

@@

\section{Strategies for detecting RDD}

@@

\section{Intelligence Collection and Agency Responsibilities}

>>>>

\section{??????? ORDERED LIST OF RECOMMENDATIONS HERE ????????}

@@

\subsection{Income}

@@

\subsection{What Makes this Different than Traditional Non-Proliferation Efforts?}

@@

\subsection{ISIL's Potential Targets (\enquote{Declaratory Policies})}

@@

This  report is directed towards the immediate response to development of the nuclear capabilities of ISIL, and does not consider the consequences which might result if ISIL were to evolve into a permanent state. This  time scale greatly reduces the possibility that ISIL possesses the infrastructure necessary to produce SNM or to assemble such material into a functioning warhead, unless they capture territory containing existing infrastructure. Due to the extreme sophistication in the triggering device, the development of thermonuclear weapons is very unlikely in the time window considered. Rather, emphasis is placed on diverting existing SNM from existing global stockpiles.

@@

\section{Limitations \& Assumptions of this Report}

@@

\autoref{app:weapons}  covers the basic categories of weapons enabled by nuclear materials: RDDs, fission weapons, and thermonuclear weapons. This Appendix builds on the concepts outlined in \autoref{app:physics} and \autoref{app:app_fuel_cycle}. Any ISIL pursuit of nuclear weapons or nuclear technology would ultimately result in the development of one of these three types of weapons.

@@

% \subsection{Introduction}

@@

\section{Post-Detonation Nuclear Forensics}

@@

\subsection{Data Processing \& Response Time}

@@

\includepdf[angle=270, pagecommand={\thispagestyle{fancy2}\null\vfill\captionof{figure}{Non-linear Logic Tree}\label{fig:logic_tree}}, scale=0.85]{./figures/NucSec_Logic_Model2.pdf}

@@

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.55]{./figures/Policy_Chart3.png}
   \caption{Summary of existing and proposed policies to combat ISIL acquisition of nuclear weapons.}
     \label{fig:policy_chart}
\end{figure}

@@

A key step for the United States in preventing ISIL acquisition and use of a nuclear device is to coordinate and organize the various initiatives across the government that are handling various pieces of this puzzle. In 2008, the authors of a Belfer Center report on preventing nuclear terrorism recommended that President Obama appoint a senior figure to a post in the National Security Council (suggesting this may be a new Deputy National Security Advisor) with the sole responsibility of waking up every morning and thinking about how to prevent a nuclear terrorist attack \cite{Bronner2008}. Since that time, President Obama created a new White House post called the White House Coordinator for Defense Policy, Countering Weapons of Mass Destruction, and Arms Control. The National Security Staff's former Senior Director for Europe, Liz Sherwood-Randall was promoted to that position \cite{Rogin2013}. While this is a step in the right direction, the arms-control role of this position is likely to consume much of Ms. Sherwood-Randall's time, forcing her to put some anti-nuclear terrorism initiatives on the backburner. For this reason, this report recommends that the White House:

\begin{itemize}
 \item Break out the countering nuclear terrorism aspect of the role of White House Coordinator for Defense Policy, Countering Weapons of Mass Destruction, and Arms Control, and create a new White House role within the National Security Council that deals exclusively with nuclear terrorism, preferable a Deputy National Security Advisor for countering nuclear terrorism. 
 \item Limit the responsibilities of this role to nuclear terrorism, leaving out nuclear weapons concerns as related to Pakistan, North Korea, and Iran, as well as arms-control. These concerns would dominate much of this person's time, detracting from the necessary focus on preventing nuclear terrorism.
 \item Fill this position with a senior figure who commands respect across cabinet departments, and is capable of speaking on behalf of the president.
\end{itemize}

This person would have full responsibility for coordinating and overseeing counter-nuclear terrorism efforts. Making this role a Deputy National Security Advisor would have the advantage of giving the role the authority to integrate policies across the Department of Defense, the Department of Energy, the State Department, the Intelligence community, Homeland Security, and law enforcement \cite{Bronner2008}. This person would also have a role working with the Office of Management and Budget (OMB) to ensure that all agency budgets are aligned with a comprehensive strategy. Additionally, the President can direct OMB to prepare a crosscut of departmental budgets to find all funds associated with preventing nuclear terrorism and put them under a single budget examiner \cite{Bronner2008}.

The above suggestions are consistent with the 2008 report from the Belfer Center. The report also emphasizes the importance of limiting the scope of this proposed role. In particular, this person should not be responsible to Pakistan, Iran, or North Korea, because these issues would take up all of this person's time \cite{Sharp2008}. This is the key problem with the current National Security Council role occupied by Ms. Sherwood-Randall - with arms control and broader counterterrorism efforts under her purview, there is still need for a senior official at this level in the White House dedicated exclusively to preventing nuclear terrorism. The full-time focus on nuclear terrorism will create an opportunity for addressing lower-level threats, as opposed to disproportionately planning for worst-case scenarios. A strategic operational plan put forward by this person can create obstacles to acquisition of nuclear weapons for terrorist groups by squeeze groups from all sides \cite{Levi2008,Ferguson2006}.  In doing this, this person may also call on collaboration from the National Counterterrorism Center.

Assuming that the White House institutionalizes this role as a Deputy National Security Advisor, the costs associated with this recommendation equal the salary and benefits of this person as well as a small staff. There will also be costs associated with the initiatives created by this person and other staff time that is devoted to those initiatives. Creating this role is an important step in crafting a strategic plan that integrates departmental efforts. The downside to this recommendation is that the efforts of this person are likely to be hampered by turf wars inherent in integrating departmental efforts; however, this difficulty can be mitigated by an approach that seeks to provide departments with a framework for evaluating and adjusting their efforts in the context of the broader effort, rather than mandating changes from the top down \cite{Levi2008}.

@@

Another White House priority should be to articulate the central importance of combating ISIL efforts to acquire and use nuclear weapons. Creating a new Deputy National Security Advisor position is the beginning of demonstrating counter-nuclear terrorism as a United States priority and broadcasting this commitment to the international community.

@@

In addition to taking the actions outlined above, the United States should launch a global security campaign aimed at building consensus that steps must be taken to reduce the risk of nuclear terrorism, encouraging allies and other nuclear powers to create high-level positions (similar to the proposed Deputy National Security Advisor), and coordinating state efforts to combat nuclear terrorism \cite{Ferguson2006}.

@@

Existing Policy: \textcolor{red}{Target ISIL Income Stream Proposed Policy}

@@

\subsubsection{Restrict ISIL Access to International Financial Systems}

@@

\subsubsection{Target ISIL Leaders \& Supporters with Sanctions}

@@

\subsection{Restrict ISIL Access to Nuclear Weapons Experts}

@@

Existing Policy: \textcolor{red}{Limiting passports, continual surveillance, and combatting propaganda}

@@

\subsection{Rollback ISIL Territorial Gains}

@@

Existing Policy: \textcolor{red}{Targets of stretgic significance in support of allies}

@@

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.5]{./figures/current_territory.png}
   \caption{Current ISIL Territory \cite{BBC2015a}}
     \label{fig:current_territory}
\end{figure}

@@

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.3]{./figures/oil_reserves.png}
   \caption{Oil Reserves and Ground Transportation Routes \cite{PoliticalCalculations2012}}
     \label{fig:oil_reserves}
\end{figure}

@@

Existing Policy: \textcolor{red}{Second Line of Defense (SLD) Core Program}

@@

In the event that ISIL attempts to acquire SNM, the core of a nuclear weapon, or a complete weapon, it is likely that it will need to be smuggled internationally. It is in the U.S.'s best interest to prevent a nuclear weapon or SNM from entering the controlled territory of ISIL. There are two methods of trafficking SNM or a complete weapon, overland and overseas. Trafficking of a complete weapon over land would be difficult due to the heavy weight of the weapon. However, the SNM is relatively light weight and volume-wise very small, and it is very likely that over land routes would be used. ISIL is extremely proficient at smuggling and makes millions of dollars a day from smuggling everything from oil to humans \cite{Amos2014}. Due to this massive illicit trafficking infrastructure already set up, it is unlikely that any effort to intercept these SNM overland would likely succeed. However, prevention of smuggling SNM overseas through the ports is much more feasible.

To prevent the smuggling of nuclear materials, the U.S., NNSA started the Second Line of Defense (SLD) Core Program  \cite{Kilmartin2010}. This program installs radiation detection equipment at borders, airports, and strategic feeder ports in Russia, former Soviet Union states, and other key countries \cite{Kilmartin2010}. While not specific to the ISIL crisis, it does help hinder any attempts for ISIL to smuggle nuclear materials/weapons. To adapt the SLD for use against ISIL, it is proposed to add SLD port monitors at the Port of Lattakia and the Port of Tartous (see \autoref{fig:seaports}), which are the closest ports to ISIL-controlled territory that have container liner services \cite{WorldPortSource2005}.  By adding two more ports to this list, the likelihood of ISIL succeeding in smuggling a nuclear weapon/materials can be dramatically reduced. In general it is estimated to cost \$3.75 million dollars per port and would take one year to implement \cite{Aloise2005}.

The NNSA is the lead agency and would be responsible for developing the strategic direction of the implementation as well as providing project performance and financial oversight. The main contributors would be the Domestic Nuclear Detection Office (DNDO), Pacific Northwest National Laboratory (PNNL), and Bechtel Nevada. The facilities at the DNDO will be used for testing purposes with the support of Bechtel Nevada. PNNL will be the lead organization for providing training to countries participating in the SLD. Bechtel Nevada will be responsible for routine and preventative maintenance for the equipment deployed overseas \cite{NationalNuclearSecurityAdministration2006}.

\begin{figure}
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.85]{./figures/seaports.png}
   \caption[Location of the five seaports in Syria \cite{WorldPortSource2014}.]{Location of the five seaports in Syria. Of these only the blue-colored ones (Port of Lattakia and the Port of Tartous) have container liner services \cite{WorldPortSource2014}.}
     \label{fig:seaports}
\end{figure}

@@

Pathway: Theft

Existing Policy:  \textcolor{red}{IAEA Monitoring}

Proposed Policy:  \textcolor{red}{Push for Additional Protocol signatures}

Lead Agency: DoS

Contributing Agency: DTRA \normalfont

The targeted capability of concern for this policy is the theft or diversion of SNM for a nuclear weapon or the theft or diversion of a complete warhead. The pathway intended for intercept is through theft or purchase, and the interception method is IAEA monitoring under the NPT Comprehensive Safeguards Agreements and, where possible, the Additional Protocol (AP). The proposal to strengthen this interception method is to negotiate with states at risk of supporting ISIL willingly or unwillingly to accept NPT AP measures to strengthen safeguards on their SNM and, if existing, any NWs the state possesses. This will help to break the links connecting “Steal” and “Purchase” on the logic tree in Figure XXX with any material needed for NW acquisition that is covered in the NPT safeguards, including “Fusion Device”, “Fission Device”, “RDD”, and “SNM”.

The lead U.S. agency responsible for implementing this policy will be the U.S. Department of State (DOS) supported by the U.S. Defense Threat Reduction Agency (DTRA). Preventing ISIL from acquiring a nuclear weapon, or at the very least SNM, falls under the purview of both the Bureau of Counterterrorism (CT) and the Bureau of International Security and Nonproliferation (ISN) within the DOS. Under this policy strategy, ISN is the principal actor in the DOS as they are the primary U.S. interface with the IAEA on the NPT [A]. DTRA is a Combat Support Agency that supports the Combatant Commanders and Services in responding to any WMD threat. One aspect of this support is managing a research and development portfolio to develop tools and capabilities to detect and respond to WMD threats. DTRA is the agency within the U.S. government that works to verify treaties related to WMD are being carried out by all signatories, and thus will be able to support the DOS ISN in this policy [B].

The AP is, “a legal document granting the IAEA complementary inspection authority to that provided in underlying safeguards agreements,” [C]. The AP’s purpose is to enable the IAEA inspectorate to provide assurances on declared and possible undeclared activities in order to have a more complete picture of a state’s overall nuclear program. The IAEA is granted increased access to information and sites under the AP. Of the 188 UN member states whom are party to the NPT, there are a total of 124 APs (plus Taiwan and Greenland) in force. Considering again Figure 2.2, the states of primary concern that are either not already party to the NPT or have not ratified the Additional Protocol include Algeria, Egypt, Sudan, South Sudan, Saudi Arabia, Yemen, Syria, and Pakistan [2, D]. Of these states, the ones with significant nuclear capabilities that are of most concern are Egypt, Syria, and Pakistan, with Pakistan being of the most concern as they are not party to the NPT and have substantial nuclear capabilities, including an existing NW stockpile [D]. DPRK is a wild card for this policy, but is unavailable through NPT and AP negotiation channels as they are not party to the NPT nor are they considered open to U.S. negotiation in general.

Therefore, this proposed policy is to negotiate with states in proximity and with ties to ISIL, particularly Egypt, Syria, and most importantly Pakistan, to accept additional IAEA NPT safeguards under the AP. Pakistan is an interesting challenge as they are not party to the NPT, but there is precedence in the case of India for strengthening IAEA NPT Safeguards through customized AP agreements. The budget for this policy is inherently small, and the schedule is longer term due to the slower mechanisms of politics.  However, it is important to pursue this policy now in order to convince the states of concern that the U.S. is open to negotiation and will not tolerate willing or unwilling support of ISIL. With a limited budget, the tools for negotiation will be so-called “carrots” and “sticks”. Carrots may include increased cooperation and development of peaceful applications of nuclear technology under the NPT regime, or increased economic trade and development with the cooperating state if nuclear energy is not a priority. Sticks may include economic sanctions, and, in more extreme cases, UN-wide sanctions.
References:
[A]	U.S. Department of State, Bureau of International Security and Nonproliferation. Nuclear Nonproliferation Treaty [Online]. Available: http:// http://www.state.gov/t/isn/npt/
[B]	Defense Threat Reduction Agency. Verifying [Online]. Available: http://http://www.dtra.mil/Missions/Verifying.aspx
[C]	International Atomic Energy Agency. (2014, October 14). IAEA Safeguards Overview: Comprehensive Safeguards Agreements and Additional Protocols [Online]. Available: https://www.iaea.org/
[D]	James Martin Center for Nonproliferation Studies. (2015, April 10). Additional Protocol [Online]. Available: http://www.nti.org

@@

\subsection{Export Control Regimes}

@@

Existing Policy: \textcolor{red}{Operation Inherent Resolve}

@@

@@

\subsection{????????  Orphan Policy Statements, need polishing and integration ????????}

@@

\chapter{Policy Courses of Action for ISIL Acquisition of a Nuclear Device}

@@

\chapter{???????  Redo captions, to have true-ordered caption citation numbers    ?????????}

\chapter{Conclusions \& Suggestions}

@@

The most substantial effect of an RDD is the spread of fear, panic, economic disruption, and the cost associated with cleaning up a contaminated area. These devices are unlikely to result in as many deaths as a fission or thermonuclear device due to the short duration of radiation exposure and the diffuse distribution of the dispersed materials. \autoref{tab:health_effects} from the U.S. Environmental Protection Agency shows the level of exposure required to produce various health symptoms and the time from acute exposure to symptom onset \cite{USEPA1999}. Department of Homeland Security (DHS) and local law enforcement crisis management plans to evacuate crowds following an RDD detonation and the subsequent radiation cleanup will drastically reduce the number of people who receive radiation doses large enough to cause short term health effects.

@@

\section{Thermonuclear}

A thermonuclear device is a multistage weapon which consists of at least one fission stage and at least one fusion stage. The fission stages are used to drive the fusion reaction, which in turn provides a short, high energy pulse of neutrons which then drives additional fission. This complicated array of triggering devices is used in turn to achieve explosive yields which far exceed a single fission device. \autoref{fig:ulam_device} shows an example of a thermonuclear device.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{./figures/Ulam_device.png}
  \caption{Schematic representation of the design of a thermonuclear weapon \cite{WikimediaCommons2005}.}
  \label{fig:ulam_device}
\end{figure}

@@

Thermonuclear weapons are far more sophisticated than fission weapons, and essentially consists of three fission devices and one fusion device. It is assumed that ISIL is pursuing nuclear weapons in order to commit acts of terrorism. While higher yields will result in larger damage,  optimizing yield does not necessarily enhance the terror factor of a nuclear device. Due to the complexity of the design of multistage thermonuclear devices, it is assumed that ISIL will elect to pursue  fission weapons, most probably a gun-type design.

@@

\section{Strategies for Detecting RDD}

@@

\section{Intelligence Collection \& Agency Responsibilities}
